---
title: "Portability from Fortran to C++"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Portability from Fortran to C++}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Origin: The Fortran ucminf Algorithm

The UCMINF optimization algorithm was originally written in **Fortran 77**
by Hans Bruun Nielsen at the Department of Mathematical Modelling,
Technical University of Denmark (DTU), in 2000. It implements a
quasi-Newton method with:

- **BFGS** (Broyden–Fletcher–Goldfarb–Shanno) inverse-Hessian updating,
- a **soft line search** that satisfies the Wolfe conditions, and
- an adaptive **trust-region** radius to safeguard against steps that are
  too large.

The original report is:

> Nielsen, H. B. (2000).
> *UCMINF — An Algorithm for Unconstrained, Nonlinear Optimization.*
> Report IMM-REP-2000-19, DTU.
> <http://www.imm.dtu.dk/documents/ftp/tr00/tr19_00.pdf>

The algorithm was later wrapped for R by Soren Hauberg and Claus Ekstrøm
in the `ucminf` CRAN package, which calls the original Fortran code via
`.Fortran()`.

## Why Reimplement in C++?

The Fortran wrapper approach has several practical limitations that
motivated the creation of `ucminfcpp`:

| Concern | Fortran + R | C++17 (ucminfcpp) |
|---------|-------------|-------------------|
| **Portability** | Requires a Fortran compiler at install time | Header-only C++17, any modern compiler |
| **Embedding** | Not straightforward without R | `#include "ucminf_core.hpp"` in any C++ project |
| **Python/Julia** | Not directly usable | `pybind11` and `CxxWrap` bindings included |
| **Maintenance** | Fixed-form Fortran 77 | Readable, idiomatic modern C++ |
| **Performance** | R ↔ Fortran call overhead | Inlinable template API; zero-allocation line search |

## The C++ Reimplementation

`ucminfcpp` rewrites the core algorithm as a **header-only C++17**
library located in `src/include/`:

```
src/
  include/
    ucminf_core.hpp        ← public API declarations
    ucminf_core_impl.hpp   ← template implementation (included by core.hpp)
  ucminf_core.c            ← low-level numeric helpers (shared with Fortran logic)
  ucminf_rcpp.cpp          ← Rcpp wrapper
```

The algorithm logic was faithfully translated line-by-line from the
Fortran source, preserving the same convergence criteria, step-size
control, and gradient tolerances. The result is **numerical equivalence**
with the original implementation.

## Technical Challenges Solved

### 1. Managing State Without Global Variables

Fortran 77 code frequently uses `COMMON` blocks for shared mutable state.
The C++ rewrite encapsulates all optimizer state inside a single `struct`
passed by reference through the call stack, eliminating global state and
making the optimizer **thread-safe** for concurrent optimization runs.

### 2. Callback from C++ into R

When used from R, the objective function is an R closure that must be
called through Rcpp. The design isolates this boundary in
`src/ucminf_rcpp.cpp`, keeping the C++ core (`ucminf_core.hpp`) free of
any R dependency — it depends only on the C++ standard library.

### 3. Zero-Allocation Line Search

The Fortran line search allocated temporary arrays on each call. The C++
version uses `std::vector` only for the iterate `x` and gradient `g`,
and reuses those buffers throughout the optimization loop, reducing heap
pressure.

### 4. Dual API for Different Use Cases

Two public entry points share the same kernel implementation:

```cpp
// High-level: accepts any callable via std::function (slight overhead)
ucminf::Result res = ucminf::minimize(x0, fn);

// Zero-overhead: compiler can inline the callable
ucminf::Result res = ucminf::minimize_direct<decltype(fn)>(x0, fn);
```

This lets downstream users choose between convenience and maximum speed.

## Verifying Equivalence

The test suite in `tests/cpp/` uses **Catch2** to verify that the C++
optimizer matches the Fortran results on several benchmark functions
(Rosenbrock, Himmelblau, sphere, etc.) to within machine epsilon.

```{r verify, eval=requireNamespace("ucminf", quietly = TRUE)}
banana <- function(x) 100 * (x[2] - x[1]^2)^2 + (1 - x[1])^2

r_cpp     <- ucminfcpp::ucminf(c(-1.2, 1), banana)
r_fortran <- ucminf::ucminf(c(-1.2, 1), banana)

stopifnot(isTRUE(all.equal(r_cpp$par, r_fortran$par)))
cat("Results are numerically identical.\n")
```
